Data Specialization for a Staged Lambda Calculus

### 2014-11-26 paper/implementation.tex
some work on the mono section

\section {Implementation and Optimization}

@@

This sort of optimization does not affect the contract of splitting at all,
because the contract only requires that there be a precomputation that the residual can accept,
with no restrictions on the form of that precomputation.
Further gains can be realized by changing the contract of splitting itself.

Let a type be {\em present} if it has no $\to$s or $\fut$s in its type.
Observe that for any $v$ with a present type, $\maskt{v}$ will always be a nested bunch of \texttt{()}s,
which is to say that it carries no information.

\nr{We probably need an example of this.}

Of course, \texttt{()} is a more compact representation that carries no information.
Thus, we can change the definition of masking so that $\maskt{v}$ is always \texttt{()},
and change the splitting rules accordingly.

Back to the boundary type problem.
The problem becomes more difficult when trying to optimize across functions, including recursive calls.
For example, consider the stage \bbtwo\ term,
\begin{lstlisting}
2`prev{
  1`letfun fact (n : int) : int = 
    if n <= 0 then 1 else fact(n-1)*n
  in next{`2`hold {`1`fact 5`2`}-100`1`}`2`
}`
\end{lstlisting}
Noting that the precomputation of the body of \texttt{fact} could be given the type,
\begin{lstlisting}
datatype prec = L | R of prec
\end{lstlisting}
the whole expression splits into:
\begin{lstlisting}
1`letfun fact (n : int) : int * prec = 
  if n <= 0 then (1,L) 
  else let (y,z) = fact(n-1) in (y*n,R z)
in fact 5`

2`l => 
letfun fact (n : unit, l0 : prec) : unit = 
  case l0 of L => () | R l1 => fact ((),l0)
in (fact (pi2 l); pi1 l)-100`
\end{lstlisting}

Observe that it's wasteful to run the stage \bbtwo\ version of \texttt{fact}, since it
always returns \texttt{()}; worse yet, it has linear runtime!
But recognizing and optimizing this away is in general a global operation,
so we instead solve this issue by adding a new staging annotation to \lang, called
\texttt{mono}. The term $\monoSt~e$ is stage \bbone, and requires that $e$ contains
no stage \bbtwo\ subexpressions. This adds a new stage \bbmono\ and two new
rules to \lang:
\begin{mathpar}
\infer{\typesone{\monoSt~e}A}{\typesmono e A & A~\mathrm{safe}} \and
\infer{A\to B~\mathrm{safe}}{A~\mathrm{present} & B~\mathrm{safe}} \and
\infer{A~\mathrm{safe}}{A~\mathrm{present}} \and
\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}} \and
\infer{\splitone{\monoSt~e}A {(e',()), \_.r} }{e \overset{C}{\rightarrow} e' & A \overset{G}{\rightarrow} r} \and
\infer{A \overset{G}{\rightarrow} ()}{A ~\mathrm{present}} \and
\infer{(A \to B)\overset{G}{\rightarrow} r}{B \overset{G}{\rightarrow} \lambda \_.r }
\end{mathpar}
Here, $\overset{C}{\rightarrow}$ translates functions to the signature expected for stage \bbone\ splitting outputs
by performing the two rewrites in place:
\begin{align*}
\lam x A e &\overset{C}{\rightarrow} \lambda x.(e,())\\
e_1~e_2 &\overset{C}{\rightarrow} \pio~(e_1~e_2)
\end{align*}
And then $\overset{G}{\rightarrow}$ generates a resumer that corresponds to the type $A$.
It operates inductively on the safety judgement of $A$.
At the safe $A$ base case, we make use of the prior optimization that says all safe $A$s have a resumer that evaluates to \texttt{()}.

>>>>

\section {Implementation and Optimization}
\label{sec:implementation}

@@

This sort of optimization does not affect the contract of splitting at all,
because the contract only requires that there be a precomputation that the residual can accept,
with no restrictions on the form of that precomputation.

The problem becomes more difficult when trying to optimize across function boundaries.
For example, consider the stage \bbtwo\ term,
\begin{lstlisting}
2`prev{
  1`letfun double (n : int) = n + n
  in next{`2`hold {`1`double 5`2`}-7`1`}`2`
}`
\end{lstlisting}
Using the rules above, this term splits into:
\begin{lstlisting}
1`letfun double (n : int) = (n+n, ())
in double 5`

2`l => 
letfun double (n : unit, l0 : unit) = ()
in (double (pi2 l); pi1 l)-100`
\end{lstlisting}

Observe that it's wasteful to run the stage \bbtwo\ version of \texttt{double}, since it always returns \texttt{()}. 
In this case, the wasted work is small, but if \texttt{double} were a recursive function,
then it's runtime may be unbounded.

There are two ways plausible ways to solve this problem:
either we could define an optimization step that comes 
after splitting and show that it tends to fix the inefficiencies that splitting creates; 
or we could give more structure to the input language so that splitting can generate more efficient code directly.

We choose to investigate the latter approach.
Specifically, we add new staging features to \lang\ which allow us to indicate that a piece of code is {\em strictly} 
stage \bbone---{\em i.e.} that it contains no $\next$ blocks.
At the type level, this manifests as a constructor called $\monoType$.
So for instance, $\rmint \to \rmint$ is a stage \bbone\ function from one integer to another,
which might fail to terminate at stage \bbtwo.  
Alternatively $\monoType (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
which {\em cannot} have any content at stage \bbtwo.
Naturally, the well-formedness of $\monoType A$ requires that $A$ contain no $\fut$s.

At the term level, we add the $\monoTerm$, $\lifttag$, \texttt{let mono} constructs to move information
in and out of $\monoType$.

\input{figures/mono.tex}
